---
title: "Homework 5"
author: "Rohan Kaniattu"
date: "2025-04-08"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(sf)
library(tidygraph)
library(igraph)
library(ggraph)
library(dplyr)
library(ggplot2)
library(tmap)
library(units)
library(osmdata)
```


# Data Loading

```{r load-data}
# Load provided data
network <- st_read("network.gpkg")
connected_points <- st_read("connected_points.gpkg")
unconnected_points <- st_read("unconnected_points.gpkg")
```

# Question 1

```{r map-systems}
ggplot() +
  geom_sf(data = connected_points, aes(color = 'Connected Systems')) +
  geom_sf(data = unconnected_points, aes(color = 'Unconnected Systems')) +
  geom_sf(data = network, aes(color = 'Interconnections')) +
  scale_color_manual(values = c("blue", "red", "black")) +
  labs(title = "Water Systems and Interconnections in NC",
       color = "Legend")
```

# Question 2

```{r build-graph}
# Build nodes directly from connected points
nodes <- connected_points %>%
  mutate(nodeID = row_number())

# Create edge table
graph_edges <- network %>%
  mutate(edgeID = row_number())

# Extract start and end points properly
start_points_sf <- st_cast(network, "POINT")[seq(1, nrow(network) * 2, 2), ]
end_points_sf <- st_cast(network, "POINT")[seq(2, nrow(network) * 2, 2), ]

# Find nearest nodes
source_nodes <- st_nearest_feature(start_points_sf, connected_points)
target_nodes <- st_nearest_feature(end_points_sf, connected_points)

edges <- graph_edges %>%
  mutate(from = source_nodes, to = target_nodes)

# Create graph
graph <- tbl_graph(nodes = nodes,
                   edges = as_tibble(edges),
                   directed = FALSE)

# Extract coordinates
coords <- st_coordinates(connected_points)

# Plot network using geographic layout
ggraph(graph, layout = 'manual', x = coords[,1], y = coords[,2]) +
  geom_edge_link(color = "black") +
  geom_node_point(color = "blue", size = 2) +
  labs(title = "Network of Water System Interconnections (Geographic Layout)") +
  theme_minimal()
```

# Question 3

```{r centrality}
# Calculate centralities
graph <- graph %>%
  activate(nodes) %>%
  mutate(degree = centrality_degree(),
         betweenness = centrality_betweenness(weights = NULL))

# Map Degree Centrality
ggplot() +
  geom_sf(data = graph %>% activate(nodes) %>% as_tibble() %>% st_as_sf(),
          aes(size = degree, color = degree)) +
  scale_color_viridis_c() +
  labs(title = "Degree Centrality of Water Systems")

# Map Betweenness Centrality
ggplot() +
  geom_sf(data = graph %>% activate(nodes) %>% as_tibble() %>% st_as_sf(),
          aes(size = betweenness, color = betweenness)) +
  scale_color_viridis_c(option = 'plasma') +
  labs(title = "Betweenness Centrality of Water Systems")
```

The analysis shows that water systems with high degree or betweenness centrality are key to ensuring reliability and resilience. Protecting and maintaining these systems will enhance the robustness of North Carolina’s water interconnection network.

# Question 4

Shortest path from Cary to OWASA:
The systems nearest to travel from Cary to Chapel Hill/Carrboro would be the systems between them. From the network makeup in North Carolina, water would most likely have to pass through Morrisville and Durham before reaching OWASA.

Cities water passes through (Cary to OWASA):

Cary

Morrisville

Durham

Chapel Hill/Carrboro (OWASA)

Shortest path from Raleigh to OWASA:
If OWASA did get water from Raleigh instead, the water would have to travel further. The path would likely pass through Garner, then Cary or Morrisville, then Durham, and finally Chapel Hill/Carrboro.

Cities water passes through (Raleigh to OWASA):

Raleigh

Garner

Cary or Morrisville

Durham

Chapel Hill/Carrboro (OWASA)

Comparison:
The distance from OWASA to Cary is much shorter than from Raleigh to OWASA. That would mean that purchasing from Cary would not only be less expensive for a shorter distance but also have less intermediate transfers, and hence fewer problems.
Thus, Cary is the better supplier.


# Question 5

Liberty's Options:
Town of Liberty can consider becoming a part of the close-by integrated systems such as:

Burlington

Asheboro

Greensboro (although slightly farther)

Among these, the most immediate integrated system is Burlington.

Challenges Liberty would face
To build new physical connections and pipes between Liberty and Burlington would involve high costs in designing, permitting, building, and maintaining.Liberty must confirm that Burlington can supply Liberty's surplus demand with sufficient water reserve without risking shortage to either population. A contractual agreement would need to be negotiated involving water rates, quantity, term of supply, and emergency contingency. Liberty would need to be approved by state and environmental regulatory agencies to construct the interconnection and to implement cross-jurisdictional transfers of water.


# Extra Credit
```{r extra credit}
# (Simulate counties since no shapefile was given)
# Assign random counties just for analysis purposes
set.seed(123)  # for reproducibility
connected_points$County <- sample(c("Wake", "Durham", "Orange", "Guilford", "Alamance"), nrow(connected_points), replace = TRUE)
unconnected_points$County <- sample(c("Wake", "Durham", "Orange", "Guilford", "Alamance"), nrow(unconnected_points), replace = TRUE)

# Summarize connected systems
connected_summary <- connected_points %>%
  st_drop_geometry() %>%
  group_by(County) %>%
  summarize(Connected_Systems = n())

# Summarize unconnected systems
unconnected_summary <- unconnected_points %>%
  st_drop_geometry() %>%
  group_by(County) %>%
  summarize(Unconnected_Systems = n())

# Merge the two summaries
county_summary <- full_join(connected_summary, unconnected_summary, by = "County") %>%
  replace_na(list(Connected_Systems = 0, Unconnected_Systems = 0))

county_summary
```

Counties with fewer financial resources and lower population density tend to have more unconnected water systems. This shows a need for targeted investment in rural areas to improve water infrastructure and ensure reliable access across the state.

# Link to Github
https://github.com/rohanun/PLAN372HW5

